---
title: "Introduction to justiceR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to justiceR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

# Introduction to justiceR

Welcome to **justiceR**, an R package that provides essential tools for criminal justice data analysis. The package is designed to simplify and enhance data-driven insights through functions that generate comprehensive codebooks, calculate key metrics, and create visualizations. As justiceR continues to evolve, more functions will be added to aid in criminal justice data analysis, making it a comprehensive resource for researchers and analysts.

This vignette will walk you through the main features of the package, focusing on the `generate_codebook()` function and demonstrating its flexibility and utility in understanding the structure of your data.

------------------------------------------------------------------------

## Installation

You can install the development version of **justiceR** from GitHub using the `devtools` package:

```{r eval=FALSE, include=FALSE}
# Install devtools if you haven't already
install.packages("devtools")

# Install justiceR from GitHub
devtools::install_github("mr4909/justiceR@develop")
```

Once installed, you can load the package:

```{r}
library(justiceR)
# Explicitly load all functions for vignette builds
devtools::load_all()
```

## Using generate_codebook()

The `generate_codebook()` function is a powerful tool for exploring and documenting your dataset. It generates a detailed codebook that provides descriptive statistics for each variable in your data frame. These statistics are particularly useful for understanding your data before performing more in-depth analyses.

**Example Data** 
To illustrate the functionality of `generate_codebook()`, let's start with a sample dataset:

```{r}
# Example data frame
df <- data.frame(
  age = c(25, 30, 22, 40, NA, 35, 28),
  start_date = as.Date(c("2020-01-01", "2020-02-15", NA, "2020-03-10", "2020-04-20", "2020-05-25", "2020-06-30")),
  active = c(TRUE, FALSE, TRUE, FALSE, TRUE, NA, FALSE),
  crime_type = factor(c("Theft", "Assault", "Theft", "Fraud", "Assault", "Theft", "Robbery"))
)
```

### Basic Usage

To generate a codebook for the dataset:

```{r}
# Generate codebook
generate_codebook(df)
```

This will create a well-organized codebook summarizing each variable:

-   **Numeric Variables:** Statistics such as minimum, average, median, maximum, and standard deviation.
-   **Categorical Variables:** The most frequent categories and their percentages.
-   **Logical Variables:** Counts and percentages of TRUE and FALSE values.
-   **Date Variables:** The earliest and latest dates in the dataset.

### Customizing Your Codebook

**Hiding Sensitive Statistics** 
For some datasets, you may want to hide statistics for variables containing personally identifiable or sensitive information. Use the hide_statistics parameter to conceal these statistics:

```{r}
# Hide statistics for the 'active' variable
generate_codebook(
  df,
  hide_statistics = c("active")
)
```

Explanation: 
The statistics for the active variable will be replaced with "Hidden" in the codebook, ensuring sensitive data is not exposed.

### Customizing the Number of Top Categories

The `generate_codebook()` function allows you to specify how many top categories should be displayed for categorical variables using the top_n parameter. By default, it shows the top 5 categories, but you can customize this to suit your needs.

```{r}
# Show only the top 2 categories for categorical variables
generate_codebook(
  df,
  top_n = 2
)
```

Explanation: 
Up to 2 top categories for each categorical variable will be displayed. If a variable has fewer than 2 categories, all categories will be shown.

### Advanced Example

Here's a more complex example demonstrating the flexibility of the `generate_codebook()` function:

```{r}
# Comprehensive data frame
df <- data.frame(
  age = c(25, 30, 22, 40, NA, 35, 28),
  start_date = as.Date(c("2020-01-01", "2020-02-15", NA, "2020-03-10", "2020-04-20", "2020-05-25", "2020-06-30")),
  active = c(TRUE, FALSE, TRUE, FALSE, TRUE, NA, FALSE),
  crime_type = factor(c("Theft", "Assault", "Theft", "Fraud", "Assault", "Theft", "Robbery"))
)

# Generate a detailed codebook
generate_codebook(
  df,
  hide_statistics = c("active"),
  top_n = 3
)
```

Explanation: 
In this example, statistics for the active variable are hidden, and only the top 3 categories are displayed for categorical variables like gender and crime_type. This level of customization allows you to tailor the codebook to your specific needs.

# Conclusion

The `generate_codebook()` function in justiceR is a versatile and powerful tool for exploring and documenting your data. It provides a detailed overview of your dataset's structure, helping you identify patterns and potential issues before deeper analysis. As justiceR continues to evolve, more functions will be added to aid in criminal justice data analysis, making it a comprehensive resource for researchers and analysts.

# Contact and Support

Developed by Mari Roberts. For questions, feedback, or to report issues, please contact [mroberts\@csg.org](mailto:mroberts@csg.org){.email}.
